Algorithmic detection of conserved quantities of finite-difference schemes for partial differential equations

### 2023-05-02 Papers/Paper ACM/tmp/CPpaper.tex
Friedmann paper

@@

Note, that the result of step 2 exists only at time $t$ because all instances of $t+1$ have been replaced. Thus, we do not need any computations of difference ideals or Gr{\"o}bner basis.
\begin{example}
	\label{timeexplicitheateqalgo}
	We check for conservation of $\sum_nu(n,t)$ under the scheme from Example \ref{timeexplicitheateq}.
	The discrete time derivative (step 1) is
	\begin{gather*}
		\sum_nu(n,t+1)-u(n,t).
	\end{gather*}
	Replacing $u(n,t+1)$ by the right-hand side of the scheme (step 2) gives
	\begin{gather*}
		\sum_nu(n,t)+(u(n+1,t)-2u(n,t)+u(n-1,t))-u(n,t).
	\end{gather*}
	Then, we compute the discrete partial variational derivative (step 3), which equals zero. Hence (step 4), we have conservation.
\end{example}
\begin{example}
	\label{heatex1}
	We verify the result from Example \ref{timeexplicitheateqalgo}, using our implementation in {\sc Mathematica} of Algorithm \ref{alg: discreteconservedqexplicit}.

@@

\subsection*{Step 1: Build the discrete time derivative}
We build the discrete time derivative by subtracting $\mathcal{J}[u](t)$ from $\mathcal{J}[u](t+1)$. This task is done in our code by the {\sc TimeDifference}.
	\begin{example}
		The discrete time derivative for $\sum_n u(n,t)$ is

@@

@@

We have implemented this algorithm in the {\sc FindDiscreteConservedQuantityBasis}.
\begin{example}
	We find a basis for conserved quantities that are generated by $u(n,t)$ and $n$ and that have at most degree 3, admitted by the scheme for the heat equation from Example \ref{heatex1}.

	\end{small}
\end{example}
\section{Examples and applications}
\label{eaa}
We illustrate our code with the Burgers equation and a system of mean-field games.
\subsection*{Burgers equation}
The Burgers equation is the PDE $u_t+uu_x=0$ \cite{smoller94} in one space dimension. Any function of $u$ is a conserved quantity. We are interested in discretizations that preserve mass $\int u\ dx$.
\begin{example}
	We check for conservation of mass using a forward-difference discretization:

	\end{small}
\end{example}
We want to find a scheme that preserves this quantity. Therefore we check for conservation under a class of schemes with a parameter, to find an appropriate choice for this parameter.
\begin{example}
	We discretize $u_x$ using a three-point stencil with a parameter $a$.

	\end{small}
	We have conservation if we choose $a=\frac{1}{2}$, corresponding to the standard central difference.
\end{example}
So far, we have only seen schemes that are in time-explicit form, but our algorithm allows also for general schemes:
\begin{example}
	We consider the scheme from the previous example but in the time-implicit version.

	\end{small}
\end{example}
\subsection*{Conserved quantities for a mean-field game}
As a second application, we use our code to study conserved quantities admitted by the discretization of a system of PDEs.
\begin{example}
	In \cite{gomes18}, Gomes et al. derived the following system of PDEs
	\begin{gather*}
		v_t+\Big(\frac{v^2}{2}-\frac{m^2}{2}\Big)_x=0\\
		m_t-(vm)_x=0.
	\end{gather*}
	This system admits the conserved quantities $\int v\ dx$ and $\int m\ dx$.
	Because this system was derived from a forward-forward mean-field game, we discretize it forward in time. We check with our code if the scheme admits the same preserved quantities as the continuous system:

	\end{small}
	We reproduce this result by noticing, that the scheme is in time-explicit form:

	\end{small}
	Those are the only conserved quantities for this scheme, that are polynomials up to degree 4 in $v$ and $m$.

	\end{small}
\end{example}
\begin{example}
	The related backward-forward system reads
	\begin{gather*}
		-v_t+\Big(\frac{v^2}{2}-\frac{m^2}{2}\Big)_x=0\\
		m_t-(vm)_x=0.
	\end{gather*}
	This system admits the same conserved quantities as the forward-forward system, but we approximate it explicitly (forward) in time for $m$ and implicitly (backward) for $v$. Our code can handle this setting, if we specify an order for the elimination of the variables, telling the code to use an explicit monomial order for $m$ and an implicit one for $v$:

>>>>

	\begin{EXE}
		LABEL="dpvd1"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"timeVars"->{t}]
		expression = u[n,t+1] - u[n,t]
		PartialDVarDOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}	

@@

Note, that the result of step 2 exists only at time $t$ because all instances of $t+1$ have been replaced. Thus, we do not need any computations of difference ideals or Gr{\"o}bner basis.
\begin{example}
	\label{timeexplicitheateqalgo}
	We check for conservation of $\sum_nu(n,t)$ under the scheme from Example \ref{timeexplicitheateq}.
	The discrete time derivative (step 1) is
	\begin{gather*}
		\sum_nu(n,t+1)-u(n,t).
	\end{gather*}
	Replacing $u(n,t+1)$ by the right-hand side of the scheme (step 2) gives
	\begin{gather*}
		\sum_nu(n,t)+(u(n+1,t)-2u(n,t)+u(n-1,t))-u(n,t).
	\end{gather*}
	Then, we compute the discrete partial variational derivative (step 3), which equals zero. Hence (step 4), we have conservation.
\end{example}
\begin{example}
	\label{heatex1}
	We verify the result from Example \ref{timeexplicitheateqalgo}, using our implementation in {\sc Mathematica} of Algorithm \ref{alg: discreteconservedqexplicit}.
	\begin{EXE}
		LABEL="rhs1"
		(*CODE*)
		variables = Association["indVars"->{n},"depVars"->{u},"rhs"->{u[n]+(u[n+1]-2*u[n]+u[n-1])}]
		expression = u[n]
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

@@

\subsection*{Step 1: Build the discrete time derivative}
We build the discrete time derivative by subtracting $\mathcal{J}[u](t)$ from $\mathcal{J}[u](t+1)$. This task is done in our code by the {\sc TimeDifference}.
	\begin{example}
		The discrete time derivative for $\sum_n u(n,t)$ is
		\begin{EXE}
			LABEL="heateq2"
			(*CODE*)
			variables = <|"indVars"->{n,t},"depVars"->{u}|>
			expression = u[n,t]
			TimeDifferenceOperator[variables][expression]
			(*CODE*)
			PUNCT={,.}
		\end{EXE}

@@

		\begin{EXE}
		LABEL="heateq10"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"scheme"->{u[n,t+1]-u[n,t]-(u[n+1,t]-2*u[n,t]+u[n-1,t])}]
		expression = u[n,t]
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
		\end{EXE}

@@

We have implemented this algorithm in the {\sc FindDiscreteConservedQuantityBasis}.
\begin{example}
	We find a basis for conserved quantities that are generated by $u(n,t)$ and $n$ and that have at most degree 3, admitted by the scheme for the heat equation from Example \ref{heatex1}.
	\begin{EXE}
		LABEL="heateq11"
		(*CODE*)
		variables = <|"indVars"->{n},"depVars"->{u},"rhs"->{u[n]+u[n+1]-2*u[n]+u[n-1]}|>
		FindDiscreteConservedQuantityBasisOperator[variables][<|"degree"->3,"generators"->{u[n],n}|>]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
\end{example}
\section{Examples and applications}
\label{eaa}
We illustrate our code with the Burgers equation and a system of mean-field games.
\subsection*{Burgers equation}
The Burgers equation is the PDE $u_t+uu_x=0$ \cite{smoller94} in one space dimension. Any function of $u$ is a conserved quantity. We are interested in discretizations that preserve mass $\int u\ dx$.
\begin{example}
	We check for conservation of mass using a forward-difference discretization:
	\begin{EXE}
		LABEL="burgers1"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"scheme"->{u[n,t+1]-u[n,t]-u[n,t]*(u[n+1,t]-u[n,t])}];
		expression = u[n,t];
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
\end{example}
We want to find a scheme that preserves this quantity. Therefore we check for conservation under a class of schemes with a parameter, to find an appropriate choice for this parameter.
\begin{example}
	We discretize $u_x$ using a three-point stencil with a parameter $a$.
	\begin{EXE}
		LABEL="burgers2"
		(*CODE*)
		variables = Association["indVars"->{n,t}, "depVars"->{u}, "pars"->{a}, "scheme"->{u[n,t+1]-u[n,t]-u[n,t]*(a*(u[n+1,t]-u[n,t])+(1-a)*(u[n,t]-u[n-1,t]))}];
		expression = u[n,t];
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
	We have conservation if we choose $a=\frac{1}{2}$, corresponding to the standard central difference.
\end{example}
So far, we have only seen schemes that are in time-explicit form, but our algorithm allows also for general schemes:
\begin{example}
	We consider the scheme from the previous example but in the time-implicit version.
	\begin{EXE}
		LABEL="burgers3"
		(*CODE*)
		variables = Association["indVars"->{n,t}, "depVars"->{u}, "pars"->{a},"elimOrder"->"implicit", "scheme"->{u[n,t+1]-u[n,t]-u[n,t+1]*(a*(u[n+1,t+1]-u[n,t+1])+(1-a)*(u[n,t+1]-u[n-1,t+1]))}]
		DiscreteConservedQOperator[variables][u[n,t]]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
\end{example}
\subsection*{Conserved quantities for a mean-field game}
As a second application, we use our code to study conserved quantities admitted by the discretization of a system of PDEs.
\begin{example}
	In \cite{gomes18}, Gomes et al. derived the following system of PDEs
	\begin{gather*}
		v_t+\Big(\frac{v^2}{2}-\frac{m^2}{2}\Big)_x=0\\
		m_t-(vm)_x=0.
	\end{gather*}
	This system admits the conserved quantities $\int v\ dx$ and $\int m\ dx$.
	Because this system was derived from a forward-forward mean-field game, we discretize it forward in time. We check with our code if the scheme admits the same preserved quantities as the continuous system:
	\begin{EXE}
		LABEL="mfg1"
		(*CODE*)
		variables = Association[
		"indVars" -> {n, t},
		"depVars" -> {v, m},
		"display result" -> True,
		"scheme" -> {v[n, t + 1] - 
			v[n, 
			t] + (v[n + 1, t]^2/2 - m[n + 1, t]^2/2) - (v[n, t]^2/2 - 
			m[n, t]^2/2), 
			m[n, t + 1] - 
			m[n, t] - (m[n + 1, t]*v[n + 1, t]) + (m[n, t]*v[n, t])}
		];
		DiscreteConservedQOperator[variables][{v[n, t], m[n, t]}]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
	We reproduce this result by noticing, that the scheme is in time-explicit form:
	\begin{EXE}
		LABEL="mfg2"
		(*CODE*)
		variables = Association[
		"indVars" -> {n},
		"depVars" -> {v, m},
		"rhs" -> { 
			v[n] -(v[n + 1]^2/2 - m[n + 1]^2/2) + (v[n]^2/2 - 
			m[n]^2/2), 
			m[n] + (m[n + 1]*v[n + 1]) - (m[n]*v[n])}
		];
		DiscreteConservedQOperator[variables][{v[n], m[n]}]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
	Those are the only conserved quantities for this scheme, that are polynomials up to degree 4 in $v$ and $m$.
	\begin{EXE}
		LABEL="mfg3"
		(*CODE*)
		variables = Association[
		"indVars" -> {n, t},
		"depVars" -> {v, m},
		"VarDOperator"->PartialDVarDOperator,
		"timeVars"->{t},
		"eqRhs" -> { 
			(v[n + 1, t]^2/2 - m[n + 1, t]^2/2) - (v[n, t]^2/2 - 
			m[n, t]^2/2), 
			(m[n + 1, t]*v[n + 1, t]) - (m[n, t]*v[n, t])}
		];
		FindConservedQuantityBasisOperator[variables][Association["degree"->4,"generators"->{v[n,t],m[n,t]}]]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}

	\end{small}
\end{example}
\begin{example}
	The related backward-forward system reads
	\begin{gather*}
		-v_t+\Big(\frac{v^2}{2}-\frac{m^2}{2}\Big)_x=0\\
		m_t-(vm)_x=0.
	\end{gather*}
	This system admits the same conserved quantities as the forward-forward system, but we approximate it explicitly (forward) in time for $m$ and implicitly (backward) for $v$. Our code can handle this setting, if we specify an order for the elimination of the variables, telling the code to use an explicit monomial order for $m$ and an implicit one for $v$:
	\begin{EXE}
		LABEL="mfg4"
		(*CODE*)
		variables = Association[
		"indVars" -> {n, t},
		"depVars" -> {m, v},
		"elimOrder" -> "explicitimplicit",
		"scheme" -> {-(v[n, t + 1] - 
			v[n, t]) + (v[n + 1, t+1]^2/2 - m[n + 1, t+1]^2/2) - (v[n, t+1]^2/2 - 
			m[n, t+1]^2/2), 
			m[n, t + 1] - 
			m[n, t] - (m[n + 1, t]*v[n + 1, t]) + (m[n, t]*v[n, t])}
		];
		DiscreteConservedQOperator[variables][{v[n, t], m[n, t]}]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
